The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth

### 2022-02-03 paper/paper.Rmd
Edited section 2.1.

Our process of cleaning builds heavily on the `tidyverse` approach [@tidyverse]. The data is first organised into "tidy data" [@WickhamHadley2014TD] and then further wrangled using the data pipeline and split-apply-combine approach [@plyr]. <!--AE comment: this is an important point that shouldn't be relegated to a parenthetical statement. --> (Tidy data shouldn't be confused with "tame data" which @tamedata coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.) The resulting (tame) data is provided in a new R package called `r yowie` which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.

@@

The NLSY79 is a longitudinal survey administered by the U.S Bureau of Labor Statistics that follows the lives of a sample of American youth born between 1957-1964 [@nlsy79]. The cohort originally included 12,686 respondents aged 14-22 when first interviewed in 1979. <!--AE comment: “It comprised of Blacks, Hispanics, economically disadvantaged non-Black non-Hispanics, and youth in the military.” Definitely needs to remove the term “Blacks” as that is considered to be offensive. Perhaps “It comprised of people sampled because they were Black, Hispanic, economically disadvantaged, or in the military.” --> It comprised of Blacks, Hispanics, economically disadvantaged non-Black non-Hispanics, and youth in the military. <!--AE comment: “1,643 members of the economically disadvantaged non-Black non-Hispanics, respectively” could perhaps be rephrased to “1,643 members who were selected because of their economic disadvantage, but were not Black or Hispanic.”-->In 1984 and 1990, two sub-samples were dropped from the interview; the dropped subjects were the 1,079 members of the military sample and 1,643 members of the economically disadvantaged non-Black non-Hispanics, respectively. Hence, 9,964 respondents remain in the eligible samples. The surveys were conducted annually from 1979 to 1994 and biennially thereafter. Data are currently available from Round 1 (1979 survey year) to Round 28 (2018 survey year).

<!--AE comment: I found the semicolons distracting. Perhaps replace them with commas and use parenthetical for sub-categories, or use a bulleted list.-->
Although the main focus area of the NLSY is labor and employment, the NLSY also covers several other topics, including education; training and achievement; household, geography, and contextual variables; dating, marriage, cohabitation; sexual activity, pregnancy, and fertility; children; income, assets and program participation; health; attitudes and expectations; and crime and substance use.

@@

<!--AE comment: first sentence can be removed or rephrased. Next sentence includes the word wages too many times. Next sentence should say “dropouts” rather than “drop outs” to be consistent with the previous sentence. -->
Now is the time to see how close we have come to the original textbook data. The original set contains wages of high school dropouts [@SingerJudithD2003Alda] from 1979 through to 1994. The refreshed data set for comparison is also wages on high school drop outs from 1979 to 2018. The original set is available in the R package `brolgar` and the refreshed data is available in the R package `r yowie`.

>>>>

Our process of cleaning builds heavily on the `tidyverse` approach [@tidyverse]. The data is first organized into "tidy data" [@WickhamHadley2014TD] and then further wrangled using the data pipeline and split-apply-combine approach [@plyr]. <!--AE comment: this is an important point that shouldn't be relegated to a parenthetical statement. --> (Tidy data shouldn't be confused with "tame data" which @tamedata coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.) The resulting (tame) data is provided in a new R package called `r yowie` which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.

@@

The NLSY79 is a longitudinal survey administered by the U.S Bureau of Labor Statistics that follows the lives of a sample of American youth born between 1957-1964 [@nlsy79]. The cohort originally included 12,686 respondents aged 14-22 when first interviewed in 1979. It comprised of people sampled because they were Blacks, Hispanics, economically disadvantaged non-Black non-Hispanics, and youth in the military. In 1984 and 1990, two sub-samples were dropped from the interview; the dropped subjects were the 1,079 members of the military sample and 1,643 members who represents economically disadvantage people, but were not Black nor Hispanics, respectively. These subsamples were dropped because they are not eligible anymore for interview [@nlsy79]. Hence, 9,964 respondents remain in the eligible samples. The surveys were conducted annually from 1979 to 1994 and biennially thereafter. Data are currently available from Round 1 (1979 survey year) to Round 28 (2018 survey year).

Although the main focus area of the NLSY is labor and employment, the NLSY also covers several other topics, including education, training, achievement, household, geography, dating, marriage, cohabitation, sexual activity, pregnancy, fertility, children, income, assets, health, attitudes and expectations, crime, and substance use.

@@

<!--AE comment: first sentence can be removed or rephrased. Next sentence includes the word wages too many times. Next sentence should say “dropouts” rather than “drop outs” to be consistent with the previous sentence. -->
The original set contains wages of high school dropouts [@SingerJudithD2003Alda] from 1979 through to 1994. The refreshed data set for comparison is also wages on high school dropouts from 1979 to 2018. The original set is available in the R package `brolgar` and the refreshed data is available in the R package `r yowie`.
